Elo World, a framework for \\ benchmarking weak chess \\ engines

### 2019-03-15 trunk/chess/papers/weak.tex
Added icons for the players and wrote about chessmaster, etc.

\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}

@@

\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\quad}
\newcommand\player[1]{\texttt{#1}}
\newcommand\deterministic{\colorbox{black}{\color{white} \textsf{det}}}
\newcommand\traditional{\colorbox{black}{\color{white} \textsf{trad}}}
% funny to use that v symbol made of leaves
\newcommand\vegetarian{\colorbox{black}{\color{white} \textsf{v}}}
\newcommand\canonical{\colorbox{black}{\color{white} \textsf{canon}}}
\newcommand\stateful{\colorbox{black}{\color{white} \textsf{state}}}
\newcommand\asymmetric{\colorbox{black}{\color{white} \textsf{asym}}}

\describeplayer{random} We must begin with the most canonical of all
strategies: Choosing a legal move uniformly at random. This is a
lovely choice for Elo World, for several reasons: It is simple to
describe. It is clearly canonical, in that anyone undertaking a
similar project would come up with the same thing. It is capable of
producing any sequence of moves, and thus completely spans the gamut
from the worst possible player to the best. If we run the tournament
long enough, it will eventually at least draw games even against a
hypothetical perfect opponent, a sort of Boltzmann Brilliancy. Note
that this strategy actually does keep state (the pseudorandom pool),
despite the admonition above. We can see this as not really state but
a simulation of an external source of ``true'' randomness. Most other
players fall back on making random moves to break ties or when their
primary strategy does not apply. \canonical

@@

\begin{figure}[ht]
\chessboard[setfen=2r4r/5n2/1p4p1/nP5p/1Ppk3P/2NPRNP1/1bRPPP2/2BQKB2 b - - 80 158,showmover=false]
\caption{\player{huddle} (white) checkmates \player{pacifist} on move
  158 with substantial help. Note that white's distal pawns have
  advanced; these are actually the same distance from the King as they
  are in their home row, since the distance metric includes diagonal
  moves. } \label{fig:huddle}
\end{figure}

@@

\begin{figure}[ht]
\chessboard[setfen=1NB1K2R/8/8/8/8/8/6p1/rnbqkbn1 b - - 19 64,showmover=false]
\caption{\player{reverse\_starting} draws against \player{reverse\_starting}
by repetition, both players having happily moved their surviving pieces into the
correct spots.} \label{fig:reversestarting}
\end{figure}

% TODO: Cyrillic tag or toast at the end
\describeplayer{cccp} Prioritize moves that Checkmate, Check, Capture
or Push, in that order. Push means to move pieces as deep as possible
into enemy territory (without any regard for their safety). Ties are
broken deterministically by the source/destination squares, so this
one is technically asymmetric. \deterministic \asymmetric

@@

% XXX include examples of symmetric players, as they are aesthetic.
\describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
when flipping vertically. Zero penalty for opposing e.g.~a \King\, with
a \king, but a small penalty when the pieces are not the same type,
and a larger penalty when they are not opposite colors. The starting
position is already symmetric this way, so this usually has the effect
of copying the opponent's moves when possible. As ``copy your
opponent's moves'' is a common (but underspecified) strategy
discovered by many children, this player is close to being canonical.
However, it admits a bit too much arbitrary choice in the penalties
assigned.

\describeplayer{sym\_mirror\_x} As \player{sym\_mirror\_y}, but
maximize symmetry when flipping horizontally. This does not make much
chess sense, but can produce aesthetic arrangements
(Figure~\ref{fig:XXX}).

% XXX show self play perhaps?
\describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
symmetry under 180 degree
% XXX latex symbol for ``degrees''
rotation of the board. An emergent priority is
to ``castle'' with the king and queen to ``fix'' them.

@@

% XXX cite blog comment?
\describeplayer{equalizer} Prefer to move a piece that has been
moved the fewest times, and then prefer moving to a square that
has been visited the fewest times. Castling counts as moving
both the king and rook, and visiting both destination squares.
This is the first strategy described that keeps meaningful state.
\canonical \stateful

@@

\describeplayer{safe} This strategy moves pieces towards squares where
they are likely to end the game alive. For this strategy and several
others, simply moving to maximize this score (e.g.~its sum or product
over all pieces) is very boring, since the score is almost always
maximized in the starting position. So this strategy actually makes
each move randomly, weighted by the total score of the resulting
positions. The scores are normalized (with the lowest-scoring move
receiving weight 0.0 and the highest 1.0) and then sampled according
to these weights. Without normalization, the play is almost identical
to uniformly random, since the weights of the resulting positions tend
to be very similar (dominated by the many pieces that don't move). But
it's pretty arbitrary.

\describeplayer{popular} Like \player{safe}, but the score for a
piece/square pair is the probability that the piece ends on that
square, whether it lives or dies. This player likes to follow the
crowd!

\describeplayer{dangerous} The dual of \player{safe}; the score is the
probability that the piece dies on that square. Note that a king is
said to die if it is checkmated or his side resigns. This player likes
to live life on the edge!

\describeplayer{rare} The dual of \player{dangerous}; the score
is one minus the probability of ending the game on that square. This
player has a thirst for adventure!

\describeplayer{survivalist} Like the above, but the score is the
ratio of the survival and death probabilities on the square. In the
data set, every piece ends alive or dead in every square (except
for the bishops, which can only legally occupy squares of their
color) at least 1000 times, so each ratio is defined. Here, the
sums of ratios have plenty of variability, and the highest
ratios are not so often on the starting squares. So with this
strategy, we simply do a weighted sample from the moves according
to their (non-normalized) scores.

\describeplayer{fatalist} The dual of \player{survivalist}; the score is
the ratio of the death and survival probability on the square. This
player knows that even if you win, you just have to keep playing over
and over again, so you might as well get it over with!

@@

% XXX citations for stockfish
\describeplayer{stockfish0} Stockfish is probably the strongest
open-source chess engine (or even publicly available engine); at full
strength its play is estimated to be around 3500 on the FIDE scale.
Aside from being quite machine dependent (it can search more moves in
a given amount of time when it has a fast CPU with many cores), there
are many options to fiddle with. Stockfish can use both opening books
and endgame tables; neither is used here. It also has a ``difficulty''
setting, which is set here to 0. Stockfish is run in a separate
process, and the board is reset before each move, but I am not
extremely hygienic about flushing e.g.~internal hash tables between
moves. One consequence of this attempt at statelessness is that
Stockfish sometimes walks into a draw by repetition in positions
where it would be able to win, because it doesn't know that it is
repeating positions.

% TODO: Does lichess publish the difficulty settings ? Nice to compare
% to my subjective skill level
\describeplayer{stockfish5} Stockfish as above, but set to difficulty
5.
\describeplayer{stockfish10} Same, at difficulty 10.
\describeplayer{stockfish15} Same, at difficulty 15.
\describeplayer{stockfish20} And difficulty 20, the maximum. Even
at this difficulty, Stockfish produces moves basically instantaneously.

\describeplayer{stockfish1m} As expected, the engine's performance
increases steadily as the difficulty setting increases (without
apparently affecting the time to make moves). I don't know what it's
doing with these settings. The true way to unleash chess engines is to
give them a lot of CPU and memory to search. Since the tournament is
run simultaneously across about 60 threads\footnote{ The computer is
  the completely egregious AMD 2990WX ``Threadripper~2'', which has 32
  cores (for 64 hardware threads) and 250 Watts of power dissipation
  at load. The torture of this CPU was part of the impetus for the
  paper.} using dozens of gigabytes of memory, and sometimes I would
play {\it Hollow Knight} while it ran, I wanted to avoid having the
performance be dependent on the scheduling environment. So here,
Stockfish is given a ``node'' budget of 1~million, hence {\sf 1m}. It
takes about one second per move when running alone, and is easily the
strongest player evaluated.

\begin{figure}[ht]
% worstfish (black) somehow beating suicide king
% white forces the mate. Note that worstfish
% does assume its opponent will pick good moves...
1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
37. Qxd7+ Bxd7++
\caption{TODO} \label{fig:worstfish}
\end{figure}

\describeplayer{worstfish} On the other hand, a strong chess engine
can also be used to play badly. When playing as white, for each legal
move, I ask Stockfish (configured as \player{stockfish0}) to evaluate
the resulting position from black's perspective.\footnote{By asking it
  to make a move, which also returns the evaluation of its move. The
  UCI protocol does not seem to offer a way to evaluate a position
  directly.} I then choose the move that produces the best position
for black. This is easily the worst player evaluated, but it is not
hard to imagine ways it could be worse. Indeed, playing for a loss is
a common twist on chess ({\it Losing
  Chess}~\cite{losingchesswikipedia}), although the rules of this
variant require that you capture a piece if you are able to (without
this constraint it is fairly boring). So, engines that support this
variant cannot be directly applied. We could use stronger settings of
Stockfish, but since it already invokes Stockfish for each legal move,
it is also one of the slowest players. Like Stockfish, it occasionally
blunders an otherwise losing position into a draw by repetition. But
most importantly, its search strategy when evaluating positions is not
applying the correct logic ($\alpha--\beta$ pruning); it assumes that
its opponent will choose strong moves, and that it will itself play
strong moves in future plies. As a result, it sometimes allows the
opponent to create a situation where \player{worstfish} is forced to
checkmate its opponent (Figure~\ref{fig:worstfish}).

\describeplayer{topple10k} Topple~\cite{topple} is another strong,
open source engine, provided to keep Stockfish on its toes. Here,
its node budget is 10,000. \traditional

\describeplayer{topple1m} Topple with a node budget of 1 million,
as \player{stockfish1m}. Stockfish almost always wins, though it
is not clear whether the budgets are actually comparable. \traditional

\describeplayer{chessmaster.nes\_lv1}
\describeplayer{chessmaster.nes\_lv2} \deterministic

@@

\describeplayer{blind\_yolo}
\describeplayer{blind\_kings}
% \describeplayer{blind\_spycheck} (described in paper, not yet run)

(I guess we can also segue by discussing handicaps)
(end with this so that we can segue into dilution methods)

\subsection{Diultion methods}
% Stockfish1m_32768 loses to reverse_starting in 3 moves!
1. e4 g5 2. Ba6 f5 3. Qh5

@@

\nocite{elo1978rating}
\nocite{topple}
// citation for knights distance:
\nocite{miller2013counting}

>>>>

% \usepackage[utf8]{inputenc}
% \usepackage[russian,english]{babel}

\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}
\usepackage{gensymb}

@@

\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\ \ }
\newcommand\player[1]{\texttt{#1}}
\newcommand\russian{
  \includegraphics[height=0.9em]{zazdarovye}
}
\newcommand\deterministic{
  \includegraphics[height=0.9em]{deterministic}
}
\newcommand\traditional{
  \includegraphics[height=0.9em]{traditional}
}
\newcommand\vegetarian{
  \includegraphics[height=0.9em]{vegetarian}
}
\newcommand\canonical{
  \includegraphics[height=0.9em]{canonical}
}
\newcommand\stateful{
  \includegraphics[height=0.9em]{stateful}
}
\newcommand\asymmetric{
  \includegraphics[height=0.9em]{asymmetric}
}

\describeplayer{random\_move} We must begin with the most canonical of all
strategies: Choosing a legal move uniformly at random. This is a
lovely choice for Elo World, for several reasons: It is simple to
describe. It is clearly canonical, in that anyone undertaking a
similar project would come up with the same thing. It is capable of
producing any sequence of moves, and thus completely spans the gamut
from the worst possible player to the best. If we run the tournament
long enough, it will eventually at least draw games even against a
hypothetical perfect opponent, a sort of Boltzmann Brilliancy. Note
that this strategy actually does keep state (the pseudorandom pool),
despite the admonition above. We can see this as not really state but
a simulation of an external source of ``true'' randomness. Most other
players fall back on making random moves to break ties or when their
primary strategy does not apply. \canonical

@@

\begin{figure}[ht]
\chessboard[setfen=2r4r/5n2/1p4p1/nP5p/1Ppk3P/2NPRNP1/1bRPPP2/2BQKB2 b - - 80 158,showmover=false]
\caption{\player{huddle} (white) checkmates \player{pacifist} on move
  158 with substantial help. Note that white's distal pawns have
  advanced; these are actually the same distance from the King as they
  would be in their home row, since the distance metric includes diagonal
  moves. } \label{fig:huddle}
\end{figure}

@@

\begin{figure}[ht]
\chessboard[setfen=1NB1K2R/8/8/8/8/8/6p1/rnbqkbn1 b - - 19 64,showmover=false]
\caption{\player{reverse\_starting} draws against itself
by repetition, both players having happily moved their surviving pieces into the
correct spots.} \label{fig:reversestarting}
\end{figure}

% TODO: Cyrillic tag or toast at the end
\describeplayer{cccp} Prioritize moves that Checkmate, Check, Capture
or Push, in that order. Push means to move pieces as deep as possible
into enemy territory (without any regard for their safety). Ties are
broken deterministically by the source/destination squares, so this
one is technically asymmetric.
% \begin{otherlanguage*}{russian}
% За здоровье!
% \end{otherlanguage*}
\deterministic \asymmetric \russian

@@

\describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
when flipping vertically. Zero penalty for opposing e.g.~a \King\, with
a \king, but a small penalty when the pieces are not the same type,
and a larger penalty when they are not opposite colors. The starting
position is already symmetric this way, so this usually has the effect
of copying the opponent's moves when possible. As ``copy your
opponent's moves'' is a common (but underspecified) strategy
discovered by many children, this player is close to being canonical.
However, it admits a bit too much arbitrary choice in the penalties
assigned.

\describeplayer{sym\_mirror\_x} As \player{sym\_mirror\_y}, but
maximize symmetry when flipping horizontally. This does not make much
chess sense, but can produce aesthetic arrangements.

\describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
symmetry under 180\degree\ rotation
of the board (Figure~\ref{fig:sym180}. An emergent priority is
to ``castle'' with the king and queen to ``fix'' them.

\begin{figure}
  \chessboard[setfen=2bkr3/8/n1r1pppp/1PppQ2P/N2BPPP1/1PPP1R1N/8/3RKB2 b - - 8 66,showmover=false]
  \caption{\player{sym\_180} (white) vs. \player{pacifist} after 66 moves. Note that the position is not quite rotationally symmetric, but is close given the material.} \label{fig:sym180}
\end{figure}

@@

\describeplayer{equalizer} Prefer to move a piece that has been
moved the fewest times, and then prefer moving to a square that
has been visited the fewest times. Castling counts as moving
both the king and rook, and visiting both destination squares.
This is the first strategy described that keeps meaningful state.
\canonical \stateful

@@

\describeplayer{safe} This strategy moves pieces towards squares where
they are likely to end the game alive. For this strategy and several
others, simply moving to maximize this score (e.g.~its sum or product
over all pieces) is very boring, since the score is almost always
maximized in the starting position. So this strategy actually makes
each move randomly, weighted by the total score of the resulting
positions. The scores are normalized (with the lowest-scoring move
receiving weight 0.0 and the highest 1.0) and then sampled according
to these weights. Without normalization, the play is almost identical
to uniformly random, since the weights of the resulting positions tend
to be very similar (dominated by the many pieces that don't move). But
it's pretty arbitrary. \stateful

\describeplayer{popular} Like \player{safe}, but the score for a
piece/square pair is the probability that the piece ends on that
square, whether it lives or dies. This player likes to follow the
crowd! \stateful

\describeplayer{dangerous} The dual of \player{safe}; the score is the
probability that the piece dies on that square. Note that a king is
said to die if it is checkmated or his side resigns. This player likes
to live life on the edge! \stateful

\describeplayer{rare} The dual of \player{dangerous}; the score
is one minus the probability of ending the game on that square. This
player has a thirst for adventure! \stateful

\describeplayer{survivalist} Like the above, but the score is the
ratio of the survival and death probabilities on the square. In the
data set, every piece ends alive or dead in every square (except
for the bishops, which can only legally occupy squares of their
color) at least 1000 times, so each ratio is defined. Here, the
sums of ratios have plenty of variability, and the highest
ratios are not so often on the starting squares. So with this
strategy, we simply do a weighted sample from the moves according
to their (non-normalized) scores. \stateful

\describeplayer{fatalist} The dual of \player{survivalist}; the score is
the ratio of the death and survival probability on the square. This
player knows that even if you win, you just have to keep playing over
and over again, so you might as well get it over with! \stateful

@@

\describeplayer{stockfish0} Stockfish~\cite{stockfish} is probably the
strongest open-source chess engine (or even publicly available
engine); at full strength its play is estimated to be around 3500 Elo
on the FIDE scale. Aside from being quite machine-dependent (it can
search more moves in a given amount of time when it has a fast CPU
with many cores), there are many options to fiddle with. Stockfish can
use both opening books and endgame tables; neither is used here. It
also has a ``difficulty'' setting, which is set here to 0. Stockfish
is run in a separate process, and the board is reset before each move,
but I am not extremely hygienic about flushing e.g.~internal hash
tables between moves. One consequence of this attempt at statelessness
is that Stockfish sometimes walks into a draw by repetition in
positions where it would be able to win, because it doesn't know that
it is repeating positions. \traditional

% TODO: Does lichess publish the difficulty settings ? Nice to compare
% to my subjective skill level
\describeplayer{stockfish5} Stockfish as above, but at difficulty 5.
\traditional

\describeplayer{stockfish10} Same, at difficulty 10.
\traditional

\describeplayer{stockfish15} Same, at difficulty 15.
\traditional

\describeplayer{stockfish20} And difficulty 20, the maximum. Even
at this difficulty, Stockfish produces moves basically instantaneously.
\traditional

\describeplayer{stockfish1m} As expected, the engine's performance
increases steadily as the difficulty setting increases (without
apparently affecting the time to make moves). I don't know what it's
doing with these settings. The true way to unleash chess engines is to
give them a lot of CPU and memory to search. Since the tournament is
run simultaneously across about 60 threads\footnote{ The computer is
  the completely egregious AMD 2990WX ``Threadripper~2,'' which has 32
  cores (for 64 hardware threads) and 250 Watts of power dissipation
  at load. The torture of this CPU was part of the impetus for the
  paper.} using dozens of gigabytes of memory, and sometimes I would
play {\it Hollow Knight} while it ran, I wanted to avoid having the
performance be dependent on the scheduling environment. So here,
Stockfish is given a ``node'' budget of 1~million, hence {\sf 1m}. It
takes about one second per move when running alone, and is easily the
strongest player (type) evaluated.
\traditional

\begin{figure}[ht]
  \chessboard[setfen=r1bK1b2/1p1pn3/p2k3N/1np1pQ2/2P2P1P/N2P4/R5B1/4B2R w - - 1 37,showmover=false]
  \caption{\player{suicide\_king} (white) to move against
    \player{worstfish} after 36 moves. Since the kings
    are already at their minimum distance, white will make
    a move at random. 37.~Qxe5++ wins for white immediately,
    but \player{suicide\_king} plays 37.~Qxd7??. The only
    legal move is 37\ldots Bxd7++, so \player{worstfish}
    must play it, and thus wins one of its only victories.}
  \label{fig:worstfish}
% full game:
% 1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
% 5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
% 9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
% 13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
% 17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
% 21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
% 25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
% 29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
% 33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
% 37. Qxd7+ Bxd7++
\end{figure}

\describeplayer{worstfish} On the other hand, a strong chess engine
can also be used to play badly. When playing as white, for each legal
move, I ask Stockfish (configured as \player{stockfish0}) to evaluate
the resulting position from black's perspective.\footnote{By asking it
  to make a move, which also returns the evaluation of its move. The
  UCI protocol does not seem to offer a way to evaluate a position
  directly.} I then choose the move that produces the best position
for black. This is easily the worst player evaluated, but it is not
hard to imagine ways it could be worse. Indeed, a common twist on
chess is to play for a loss, called {\it Antichess} or {\it Losing
  Chess}~\cite{wikipedialosingchess}. Recently it was even proved that
white can always win (i.e.~lose)~\cite{watkins2017losing} in this
variant! However, the variant requires that you capture a piece if
you are able to, so strategies and engines that support this variant
cannot be directly applied.
We could use stronger settings of Stockfish, but since it already
invokes Stockfish for each legal move, it is also one of the slowest
players. Like Stockfish, it occasionally blunders an otherwise losing
position into a draw by repetition. But most importantly, its search
strategy when evaluating positions is not applying the correct logic
($\alpha$--$\beta$ pruning); it assumes that its opponent will choose
strong moves, and that it will itself play strong moves in future
plies. As a result, it sometimes allows the opponent to create a
situation where \player{worstfish} is forced to checkmate its opponent
(Figure~\ref{fig:worstfish}).

\describeplayer{topple10k} Topple~\cite{topple} is another strong
open source engine, provided to keep Stockfish on its toes. Here,
its node budget is 10,000. \traditional

\describeplayer{topple1m} Topple with a node budget of 1 million,
which like \player{stockfish1m} takes about one second per move.
Stockfish almost always wins, though it is not clear whether the
budgets are actually comparable. \traditional

\describeplayer{chessmaster.nes\_lv1} This is {\it Chessmaster} for
the Nintendo Entertainment System, published in AD 1989. Moves are
extracted from the game via emulation. This proved to be somewhat
delicate, because the in-memory representation of the board is not
that simple (it appears to be represented in two parallel arrays,
perhaps using the ``0x88 method'') and the game understandably goes
wild if you mess it up. To play a move, I restore a saved machine
state in the ``board editor'' mode, and then modify memory to contain
the desired board. I then emulate button presses to operate the game
menu and return to ``playing'' mode. Chessmaster repairs its internal
data structures from this board, and makes a move. During this time I
mash controller buttons to skip its dramatic tunes and modal messages
like \verb+CHECK+. Once some memory locations reach certain values,
the move has been played; I can diff the before and after boards to
uniquely determine the move. Since this approach uses emulation, it
would normally be completely deterministic, but I deliberately stall
for a random number of frames so that it can advance its internal
pseudorandom state. A nice thing about this engine is that it is an
earnest attempt at writing a good engine, but limited to the
techniques of the 1980s, and running on hardware that was underpowered
even for its day. It finishes well behind the modern engines, but
ahead of all the non-traditional strategies. \traditional

\describeplayer{chessmaster.nes\_lv2} Same as
\player{chessmaster.nes\_lv1}, but increasing the ``Level Of Play'' to
``Newcomer 2.'' Chessmaster has stronger levels still, but in these
modes it will think for several NES minutes, which takes multiple
seconds to emulate---too slow for our purposes. Still much
weaker than modern engines (Figure~\ref{fig:chessmaster}).
\traditional

\begin{figure}[ht]
  \includegraphics[width=0.75 \linewidth]{chessmaster}
  \caption{\player{stockfish1m} (2019; running on a modern machine) beats
    \player{chessmaster.nes\_lv2} (1989; running on an emulated NES)
    thousands of times without a loss or draw.} \label{fig:chessmaster}
\end{figure}

@@

\describeplayer{blind\_yolo} This player is only allowed to see the
64-bit mask of where pieces are placed on the board, but not their
colors or types. It is described in {\it Color- and Piece-blind
  Chess}~\cite{blind}.

\describeplayer{blind\_kings} As \player{blind\_yolo}, but forcing
the prediction to produce exactly one king for each side, which
improves its accuracy a little.

\describeplayer{blind\_spycheck} As \player{blind\_kings}, but
performing a ``spy check'' before each move. Here, the player tries
capturing each piece of a given (predicted) color with other pieces
of that same (predicted) color. If such a move is legal, then one
of the two pieces was mispredicted, so we prefer the capture over
sending an incorrect position to the engine.

Each of these uses a neural network to predict the configuration of
the board, and then the equivalent of \player{stockfish1m} to make a
move for that predicted board. (If that move is illegal because the
board was mispredicted, then it plays a random legal move.) These
players can therefore be seen as handicaps on \player{stockfish1m}.

\subsection{Diultion methods}

Even at its weakest setting, \player{stockfish0} crushes all of the
nontraditional strategies. This is not surprising, but it creates
a problem for quantifying their difference in skill (do they win
one in a million games? One in 10^{100}?). Having intermediate
players allows some Elo points to flow between the two tiers
transitively. One nice way to construct intermediate players is
by interpolation. We can interpolate between any two players,\footnote{
  Even if they are stateful! The interface separates ``please
  give me a move for the current state'' and ``the following
  move was played; please update your state,'' allowing them
  to disagree. So we can just keep two independent states for
  the two players.} but interpolating with the most canonical
player, \player{random\_move}.

% Stockfish1m_32768 loses to reverse_starting in 3 moves!
1. e4 g5 2. Ba6 f5 3. Qh5

@@

\nocite{elo1978rating}
\nocite{topple}
// citation for knights distance:
\nocite{miller2013counting}
\nocite{watkins2017losing}
